---
title: "Cyclistic bike-share analysis case study"
output: html_document
date: "2022-11-17"
---
This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman (found here: <https://artscience.blog/home/divvy-dataviz-case-study>). The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: "In what ways do members and casual riders use Divvy bikes differently?"

# Install required packages

* **tidyverse** for data import and wrangling
* **lubridate** for date functions
* **ggplot** for visualization

```{r setup, eval=FALSE}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("skimr")
```

# load packages

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(skimr)
library(scales)
```

## display your working directory

```{r}
getwd()
```
## set your working directory 

```{r}
setwd("~/R/PROJECT/case1")
```

***
# STEP 1: COLLECT DATA

Uploading csv files

```{r}
d1 <- read.csv("2022_01.csv")
d2 <- read.csv("2022_02.csv")
d3 <- read.csv("2022_03.csv")
d4 <- read.csv("2022_04.csv")
d5 <- read.csv("2022_05.csv")
d6 <- read.csv("2022_06.csv")
d7 <- read.csv("2022_07.csv")
d8 <- read.csv("2022_08.csv")
d9 <- read.csv("2022_09.csv")
d10 <- read.csv("2022_10.csv")
```

# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

Compare column names each of the files
While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r}
colnames(d1)
colnames(d2)
colnames(d3)
colnames(d4)
colnames(d5)
colnames(d6)
colnames(d7)
colnames(d8)
colnames(d9)
colnames(d10)
```
All the name are same.
Now inspect the dataframes and look for incongruencies

```{r}
str(d1)
str(d2)
str(d3)
str(d4)
str(d5)
str(d6)
str(d7)
str(d8)
str(d9)
str(d10)
```

Stack individual data frames into one big data frame

```{r}
all_trips <- bind_rows(d1,d2,d3,d4,d5,d6,d7,d8,d9,d10)

```

Remove lat & lng fields

```{r}
all_trips <- all_trips[, !(names(all_trips) %in% c("start_lat", "start_lng", "end_lat", "end_lng"))]

# or use this 
# all_trips <- all_trips %>% select(-c(start_lat,start_lng,end_lat,end_lng))
```


***

# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

* Inspect the new table that has been created

```{r}
colnames(all_trips)
```

* How many rows are in data frame?

```{r}
nrow(all_trips)
```
* Dimensions of the data frame?

```{r}
dim(all_trips)
```
* See the first 6 rows of data frame

```{r}
head(all_trips)
```

* Statistical summary of data

```{r}
summary(all_trips)
```


the columns started_at and ended_at are in character object, we want to convert it into data/time [using as.POSIXlt()]
This will allow us to aggregate ride data for each month, day, or year.


```{r}
all_trips$started_at <- as.POSIXlt(all_trips$started_at , format="%Y-%m-%d %H:%M:%S" ,tz = "America/Chicago")
all_trips$ended_at <- as.POSIXlt(all_trips$ended_at , format="%Y-%m-%d %H:%M:%S" , tz = "America/Chicago")
```

 
Adding columns that list the date, month, day, and year of each ride

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Adding a "ride_length" calculation to all_trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

* Inspect the structure of the columns

```{r}
str(all_trips)
```

* Remove "bad" data
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

We will create a new version of the dataframe (v2) since data is being removed

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

* Descriptive analysis on ride_length (all figures in seconds)

```{r}
all_trips_v2 %>% 
  summarize(ride_mean = mean(ride_length),
            ride_median = median(ride_length),
            ride_max = max(ride_length),
            ride_min = min(ride_length))
```

### Compare members and casual users
* mean -(members and casual)

```{r}
aggregate(ride_length ~ member_casual, all_trips_v2, FUN = mean)
```

* median -(members and casual)

```{r}
aggregate(ride_length ~ member_casual, all_trips_v2, FUN = median)
```

* max -(members and casual)

```{r}
aggregate(ride_length ~ member_casual, all_trips_v2, FUN = max)
```

* min -(members and casual)

```{r}
aggregate(ride_length ~ member_casual, all_trips_v2, FUN = min)
```

* the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

the days of the week are out of order. 

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

let's run the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

* analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)	
  

```
# visualize

* Let's visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma)


```
* Let's create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

Creating a csv file so we can visualize in Excel or Tableau



```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

exporting counts data frame to csv

```{r}
write.csv(counts, "C:\\Users\\sonny\\Documents\\R\\PROJECT\\case1\\counts_rides.csv", row.names= FALSE)
```


